\vspace{-.5cm} Lights: a generalized joint model for high-dimensional multivariate longitudinal data and censored durations \vspace{.5cm}

### 2022-03-16 manuscript/paper.tex
minor update

Now in the case $g = g_k$, one has
\begin{equation}
\label{eq:hat_pi_ik-def}
  \E_{\theta^{(w)}}[g_k(Gi) | t_i, \delta_i, y_i] = \pi_{ik}^{\theta^{(w)}} = \dfrac{\pi_{\xi_k^{(w)}}(x_i) \Lambda_{ik,1}^{\theta^{(w)}}}{\sum_{k=0}^{K-1} \pi_{\xi_k^{(w)}}(x_i) \Lambda_{ik,1}^{\theta^{(w)}}}.
\end{equation}

\paragraph*{Proximal Quasi-Newton M-step.}
Here, we need to compute \[\theta^{(w+1)} \in \argmin_{\theta \in \R^\vartheta} \cQ_n(\theta, \theta^{(w)}) + \sum_{k=0}^{K-1} \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta} + \zeta_{2,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}.\]
We then present the parameters updates in the order given in Algorithm~\ref{alg:prox-QNEM}.
First, the update of $D^{(w)}$ is naturally given in closed-form by
\begin{equation}
  \label{eq:update_D}
  D^{(w+1)} = n^{-1} \sum_{i=1}^n \E_{\theta^{(w)}}[ b_i b_i^\top | t_i, \delta_i, y_i].
\end{equation}
Then, let us focus on the update of $\xi_k^{(w)}$ for $k=0, \ldots, K-1$. We denote
\[ P^{(w)}_{n, k}(\xi_k) = -n^{-1} \sum_{i=1}^n \sum_{k=0}^{K-1} \pi_{ik}^{\theta^{(w)}} \log \pi_{\xi_k}(x_i) \]
based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ that depend on $\xi_k$. The update for $\xi_k^{(w)}$ therefore requires to solve the following convex minimization problem
\begin{equation}
  \label{eq:minimization-xi_k}
  \xi_k^{(w+1)} \in \argmin_{\xi_k \in \R^p} P^{(w)}_{n,k}(\xi_k) + \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta}.
\end{equation}
It looks like the logistic regression objective, where labels are not fixed but softly encoded by the expectation step (computation of $\pi_{ik}^{\theta^{(w)}}$ in~\eqref{eq:hat_pi_ik-def}). We then choose to solve~\eqref{eq:minimization-xi_k} using the L-BFGS-B algorithm~\citep{zhu1997algorithm} which belongs to the class of quasi-Newton optimization routines and solves the given minimization problem by computing approximations
of the inverse Hessian matrix of the objective function. It can deal with differentiable convex objectives with box constraints.
In order to use it with $\ell_1$ penalization, which is not differentiable, we use the trick borrowed from \citet{andrew2007scalable}: for $a \in \R$, write $|a| = a^+ + a^-$, where $a^+$ and $a^-$ are respectively the positive and negative part of $a$, and add the constraints $a^+ \geq 0$ and $a^- \geq 0$.
Namely, we rewrite the minimization problem~\eqref{eq:minimization-xi_k} as the following differentiable problem with box constraints
\begin{equation}
  \label{eq:sub-problem-xi_k}
  \begin{split}
    \text{minimize}& \quad \quad P_{n, k}^{(w)}(\xi_k^+ - \xi_k^-) + \zeta_{1,k} \big((1 - \eta) \sum_{j=1}^p (\xi_{k,j}^+ + \xi_{k,j}^-) + \dfrac \eta 2 \norm{{\xi_k}^+ - {\xi_k}^-}_2^2 \big) \\
    \text{subject to}& \quad \quad \xi_{k,j}^+ \geq 0 \text{ and } \xi_{k,j}^- \geq 0 \text{ for } j = 1, \ldots, p
  \end{split} 
\end{equation}
where $\xi_k^\pm = (\xi_{k,1}^\pm, \ldots, \xi_{k,p}^\pm)^\top$. The L-BFGS-B solver requires the exact value of the gradient, which is easily given by
\begin{equation}
  \label{eq:grad-xi_k}
  \frac{\partial P_{n,k}^{(w)}(\xi_k)}{\partial \xi_k} = -n^{-1} \sum_{i=1}^n  \big(\pi_{ik}^{\theta^{(w)}} - \pi_{\xi_k}(x_i) \big) x_i.
\end{equation}
In practice, we use the \texttt{Python} solver \texttt{fmin\_l\_bfgs\_b} from \texttt{scipy.optimize}~\citep{virtanen2020scipy}. Note that for computational reasons, it is not clever to use a too small solver tolerance in the early steps of the overall algorithm, where we call tolerance the value for which iterations stop when the stopping criterion is below it. To speed up the overall algorithm, we then decrease the L-BFGS-B tolerance with the iterations $w$.

@@

\begin{algorithm}[!htb]
\DontPrintSemicolon
\caption{prox-QNEM algorithm for the lights model inference}
\label{alg:prox-QNEM}
\KwInput{Training data $\cD_n$; tuning hyper-parameters $(\zeta_{1,k}, \zeta_{2,k})_{k=0,\ldots,K-1}$}
\KwOutput{Last parameters $\hat \theta \in \R^\vartheta$}
\tcc{Initialization}
Compute the starting parameters $\theta^{(0)} \in \R^\vartheta$ \\
\For{$w = 1,\ldots,$ until convergence}    
{ 
  \tcc{E-step}
  Compute $\Lambda_{ik,\tilde g}^{\theta^{(w)}}$ using~\eqref{eq:Lambda_ik-b_i} for $\tilde g_1$ and $\tilde g_2$\\
  Compute $\pi_{ik}^{\theta^{(w)}}$ using~\eqref{eq:hat_pi_ik-def}\\
  \tcc{Proximal Quasi-Newton M-step}
  Update $D^{(w+1)}$ using~\eqref{eq:update_D}\\
  Update $(\xi_k^{(w+1)})_{k=0,\ldots,K-1}$ by solving~\eqref{eq:sub-problem-xi_k} \\
  Update $(\beta_k^{(w+1)})_{k=0,\ldots,K-1}$ by solving~\eqref{eq:update_beta} \\
  Update $(\gamma_k^{(w+1)})_{k=0,\ldots,K-1}$ by solving~\eqref{eq:gamma-update} \\
  Update $\lambda_0^{(w+1)}$ using~\eqref{eq:update_lambda_0}\\
  Update $\phi^{(w+1)}$ using~\eqref{eq:update_phi} \\
}
\textbf{Return:} {$\hat \theta$}
\end{algorithm}

@@

The hyper-parameters $(\zeta_1, \zeta_2)^\top \in \R_+^2$ are then tuned using a Bayesian optimization method~\citep{snoek2012practical}, with predictive marker defined in Section~\ref{sec:evaluation strategy} and the C-index score defined in Section~\ref{sec:Metrics}. In practice, we use the snoek2012practical\texttt{Python} package \texttt{Hyperopt}~\citep{bergstra2013hyperopt} and the Tree-structured Parzen Estimator~\citep{bergstra2011algorithms}, which is a sequential model-based optimization (SMBO) approach. SMBO methods sequentially construct models to approximate the performance of hyperparameters based on historical measurements, and then subsequently choose new hyperparameters to test based on this model.
Moreover, the search interval for hyper-parameter $\zeta_1$ is automatically computed in a data driven way described in Appendix~\ref{sec:auto-grid-cv}, while the search intervals for $\zeta_2$ is determined empirically.
% TODO: Figure de visualisation de la CV de (\zeta_1, \zeta_2) en 3D

@@

posterior risk classification for subject $i$ and latent class $k$ can be made through $\pi_{ik}^{\hat\theta}$ (see~\eqref{eq:hat_pi_ik-def}). Similar posterior probabilities have been considered for goodness-of-fit evaluation in other JLCMs models, see~\citet{proust2014joint} for instance.

@@

where $\hat S$ is obtained with Algorithm~\ref{alg:construct-set-S} using $\hat D$ (that is $S^{(w_{max})}$ for the last iteration $w_{max}$), and the density $f$ in the rigth hand side of the previous equation being defined in~\eqref{eq:ind-hyp}.
Note that marker~\eqref{eq:marker-def} is in fact obtained from the definition of $\hat \pi_{ik}^{\hat\theta}$ in~\eqref{eq:hat_pi_ik-def} taking $(t_i, \delta_i) = (t^{max}_i, 0)$, and turns out to be an estimate of the following posterior probability 
\[\P_\theta[G_i=k | T^\star_i > t^{max}_i, y_i].\]

@@

Note that other metrics have been proposed to compare individual risk predictions. In~\citet{blanche2015quantifying} for instance, authors use dynamic prediction accuracy curves (of AUC or Brier score) with the idea of moving what is called the ``landmark time'', that is the time at which predictions are made and on which the amount of available information depends, corresponding to our $t^{max}$. But comparing curves in an automatic way (in a cross-validation process for instance) is never easy, plus the comparisons proposed in the aforementioned paper require \textit{a priori} choices of time window thresholds (in particular to get back to binary predictions, being always questionable in survival analysis~\citep{bussy2019comparison}), which is not the case of the C-index metric.
Nevertheless, the idea of chosing $t^{max}$ when it comes to simulate data for models performances comparisons will be discussed in Section~\ref{sec:simulation study}, and it is an interesting notion to consider since it can be sensed that the closer $t^{max}$ gets to $T^\star$, the easier it will be for a model to make good predictions.

>>>>

Now in the case $g = g_k$, one has
\begin{equation}
\label{eq:pi_ik-def}
  \E_{\theta^{(w)}}[g_k(Gi) | t_i, \delta_i, y_i] = \pi_{ik}^{\theta^{(w)}} = \dfrac{\pi_{\xi_k^{(w)}}(x_i) \Lambda_{ik,1}^{\theta^{(w)}}}{\sum_{k=0}^{K-1} \pi_{\xi_k^{(w)}}(x_i) \Lambda_{ik,1}^{\theta^{(w)}}}.
\end{equation}

\paragraph*{Proximal Quasi-Newton M-step.}
Here, we need to compute \[\theta^{(w+1)} \in \argmin_{\theta \in \R^\vartheta} \cQ_n(\theta, \theta^{(w)}) + \sum_{k=0}^{K-1} \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta} + \zeta_{2,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}.\]
We then present the parameters updates in the order given in Algorithm~\ref{alg:prox-QNEM}.
First, the update of $D^{(w)}$ is naturally given in closed-form by
\begin{equation}
  \label{eq:update_D}
  D^{(w+1)} = n^{-1} \sum_{i=1}^n \E_{\theta^{(w)}}[ b_i b_i^\top | t_i, \delta_i, y_i].
\end{equation}
Then, let us focus on the update of $\xi_k^{(w)}$ for $k=0, \ldots, K-1$. We denote
\[ P^{(w)}_{n, k}(\xi_k) = -n^{-1} \sum_{i=1}^n \sum_{k=0}^{K-1} \pi_{ik}^{\theta^{(w)}} \log \pi_{\xi_k}(x_i) \]
based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ that depend on $\xi_k$. The update for $\xi_k^{(w)}$ therefore requires to solve the following convex minimization problem
\begin{equation}
  \label{eq:minimization-xi_k}
  \xi_k^{(w+1)} \in \argmin_{\xi_k \in \R^p} P^{(w)}_{n,k}(\xi_k) + \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta}.
\end{equation}
It looks like the logistic regression objective, where labels are not fixed but softly encoded by the expectation step (computation of $\pi_{ik}^{\theta^{(w)}}$ in~\eqref{eq:pi_ik-def}). We then choose to solve~\eqref{eq:minimization-xi_k} using the L-BFGS-B algorithm~\citep{zhu1997algorithm} which belongs to the class of quasi-Newton optimization routines and solves the given minimization problem by computing approximations
of the inverse Hessian matrix of the objective function. It can deal with differentiable convex objectives with box constraints.
In order to use it with $\ell_1$ penalization, which is not differentiable, we use the trick borrowed from \citet{andrew2007scalable}: for $a \in \R$, write $|a| = a^+ + a^-$, where $a^+$ and $a^-$ are respectively the positive and negative part of $a$, and add the constraints $a^+ \geq 0$ and $a^- \geq 0$.
Namely, we rewrite the minimization problem~\eqref{eq:minimization-xi_k} as the following differentiable problem with box constraints
\begin{equation}
  \label{eq:sub-problem-xi_k}
  \begin{split}
    \text{minimize}& \quad \quad P_{n, k}^{(w)}(\xi_k^+ - \xi_k^-) + \zeta_{1,k} \big((1 - \eta) \sum_{j=1}^p (\xi_{k,j}^+ + \xi_{k,j}^-) + \dfrac \eta 2 \norm{{\xi_k}^+ - {\xi_k}^-}_2^2 \big) \\
    \text{subject to}& \quad \quad \xi_{k,j}^+ \geq 0 \text{ and } \xi_{k,j}^- \geq 0 \text{ for } j = 1, \ldots, p
  \end{split} 
\end{equation}
where $\xi_k^\pm = (\xi_{k,1}^\pm, \ldots, \xi_{k,p}^\pm)^\top$. The L-BFGS-B solver requires the exact value of the gradient, which is easily given by
\begin{equation}
  \label{eq:grad-xi_k}
  \frac{\partial P_{n,k}^{(w)}(\xi_k)}{\partial \xi_k} = -n^{-1} \sum_{i=1}^n  \big(\pi_{ik}^{\theta^{(w)}} - \pi_{\xi_k}(x_i) \big) x_i.
\end{equation}
In practice, we use the \texttt{Python} solver \texttt{fmin\_l\_bfgs\_b} from \texttt{scipy.optimize}~\citep{virtanen2020scipy}. Note that for computational reasons, it is not clever to use a too small solver tolerance in the early steps of the overall algorithm, where we call tolerance the value for which iterations stop when the stopping criterion is below it. To speed up the overall algorithm, we then decrease the L-BFGS-B tolerance with the iterations $w$.

@@

\begin{algorithm}[!htb]
\DontPrintSemicolon
\caption{prox-QNEM algorithm for the lights model inference}
\label{alg:prox-QNEM}
\KwInput{Training data $\cD_n$; tuning hyper-parameters $(\zeta_{1,k}, \zeta_{2,k})_{k=0,\ldots,K-1}$}
\KwOutput{Last parameters $\hat \theta \in \R^\vartheta$}
\tcc{Initialization}
Compute the starting parameters $\theta^{(0)} \in \R^\vartheta$ \\
\For{$w = 1,\ldots,$ until convergence}    
{ 
  \tcc{E-step}
  Compute $\Lambda_{ik,\tilde g}^{\theta^{(w)}}$ using~\eqref{eq:Lambda_ik-b_i} for $\tilde g_1$ and $\tilde g_2$\\
  Compute $\pi_{ik}^{\theta^{(w)}}$ using~\eqref{eq:pi_ik-def}\\
  \tcc{Proximal Quasi-Newton M-step}
  Update $D^{(w+1)}$ using~\eqref{eq:update_D}\\
  Update $(\xi_k^{(w+1)})_{k=0,\ldots,K-1}$ by solving~\eqref{eq:sub-problem-xi_k} \\
  Update $(\beta_k^{(w+1)})_{k=0,\ldots,K-1}$ by solving~\eqref{eq:update_beta} \\
  Update $(\gamma_k^{(w+1)})_{k=0,\ldots,K-1}$ by solving~\eqref{eq:gamma-update} \\
  Update $\lambda_0^{(w+1)}$ using~\eqref{eq:update_lambda_0}\\
  Update $\phi^{(w+1)}$ using~\eqref{eq:update_phi} \\
}
\textbf{Return:} {$\hat \theta$}
\end{algorithm}

@@

The hyper-parameters $(\zeta_1, \zeta_2)^\top \in \R_+^2$ are then tuned using a Bayesian optimization method~\citep{snoek2012practical}, with predictive marker defined in Section~\ref{sec:evaluation strategy} and the C-index score defined in Section~\ref{sec:Metrics}. In practice, we use the \texttt{Python} package \texttt{Hyperopt}~\citep{bergstra2013hyperopt} and the Tree-structured Parzen Estimator~\citep{bergstra2011algorithms}, which is a sequential model-based optimization (SMBO) approach. SMBO methods sequentially construct models to approximate the performance of hyperparameters based on historical measurements, and then subsequently choose new hyperparameters to test based on this model.
Moreover, the search interval for hyper-parameter $\zeta_1$ is automatically computed in a data driven way described in Appendix~\ref{sec:auto-grid-cv}, while the search intervals for $\zeta_2$ is determined empirically.
% TODO: Figure de visualisation de la CV de (\zeta_1, \zeta_2) en 3D

@@

posterior risk classification for subject $i$ and latent class $k$ can be made through $\pi_{ik}^{\hat\theta}$ (see~\eqref{eq:pi_ik-def}). Similar posterior probabilities have been considered for goodness-of-fit evaluation in other JLCMs models, see~\citet{proust2014joint} for instance.

@@

where $\hat S$ is obtained with Algorithm~\ref{alg:construct-set-S} using $\hat D$ (that is $S^{(w_{max})}$ for the last iteration $w_{max}$), and the density $f$ in the rigth hand side of the previous equation being defined in~\eqref{eq:ind-hyp}.
Note that marker~\eqref{eq:marker-def} is in fact obtained from the definition of $\hat \pi_{ik}^{\hat\theta}$ in~\eqref{eq:pi_ik-def} taking $(t_i, \delta_i) = (t^{max}_i, 0)$, and turns out to be an estimate of the following posterior probability 
\[\P_\theta[G_i=k | T^\star_i > t^{max}_i, y_i].\]

@@

Note that other metrics have been proposed to compare individual risk predictions. In~\citet{blanche2015quantifying} for instance, authors use dynamic prediction accuracy curves (of AUC or Brier score) with the idea of moving what is called the ``landmark time'', that is the time at which predictions are made and on which the amount of available information depends, corresponding to our $t^{max}$. But comparing curves in an automatic way (in a cross-validation process for instance) is never easy, plus the comparisons proposed in the aforementioned paper require \textit{a priori} choices of time window thresholds (in particular to get back to binary predictions, being always questionable in survival analysis~-v\citep{bussy2019comparison}), which is not the case of the C-index metric.
Nevertheless, the idea of chosing $t^{max}$ when it comes to simulate data for models performances comparisons will be discussed in Section~\ref{sec:simulation study}, and it is an interesting notion to consider since it can be sensed that the closer $t^{max}$ gets to $T^\star$, the easier it will be for a model to make good predictions.
